Improving the prospects for multimessenger astronomy with \earlywarning\ detection of compact binary coalescence.

### 2011-05-27 method.tex
misc section 2 cleanups and rewordings
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@761 ccdd05d5-56c1-461a-bf98-85f6876523b4

In this section we describe a decomposition of the compact binary parameter
space that reduces low latency filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multiband decomposition of the compact binary parameter space that resulted in
search with $\sim$minutes latency in \LIGO's S6 and Virgo's VSR2 science runs.
We combine this with the orthogonal decomposition described
in~\cite{Cannon:2010p10398} that exploits the redundancy of the template banks.

@@

Inspiral signals are parameterized by a set of intrinsic parameters
$\bar{\theta}$ that determine the amplitude and phase evolution of the observed
binary signals.  Searches for gravitational waves from compact binary
coalescence typically employ matched filter banks, called template banks,
\cite{findchirppaper} that discretely sample the possible intrinsic parameters.
The filter models of the waveforms, known as templates are weighted by the
detector noise amplitude spectral density, $x(t,\bar{\theta})$ and filtered
against the whitened detector data.  A sufficient number of templates are
chosen to assure a minimal loss of \SNR~\cite{Owen:1995tm,Owen:1998dk}.  To
construct a template bank, matched filters are chosen with discrete signal
parameters $\theta_1,\, \theta_2,\, \cdots$, $\theta_N$, such that any possible
signal will have a cross-correlation of $\geq0.97$ with at least one template.
Such a template bank is said to have a 97\% {\em minimum match}. For systems
where the effects of spin can be ignored, the intrinsic parameters are the
component masses of the binary, $\bar{\theta} = (m_1, m_2)$.

The procedure for choosing a bank of matched filters for gravitational wave
searches is known~\cite{Owen:1995tm}. Therefore, here we will assume that a
bank of filters has already been chosen based on the targetted parameter space
and we will denote the $i^{\mathrm th}$ filter with parameters $\bar{\theta_i}$
as a function of time $x_i(\tau)$.%
%
\footnote{There are two gravitational wave polarizations, $+$ and $\times$. A
given detector will observe a combination of these polarizations that will
largely be degenerate with an overall unknown constant phase.  This can be
maximized over by filtering for quadrature phases and taking the magnitude of
the result. For simplicity we will ignore that aspect in this work as it is
straightforward to generalize, but not necessary for understanding any of the
points that will be made.}
%
In this work we will discuss transformations to a set of filters
$\{x_i(\tau)\}$.  Some of these transformations are not useful or practical
over the entire parameter space.  For that reason we assume from here onward
that the set of filters $\{x_i(\tau)\}$ refer to a set of near neighbor filters
that can be chosen as a subset of the full parameter space.  Several such
subsets can be chosen until all of the filters are a member of one local set.

@@

The evaluation of the integrals in \eqref{eq:SNRTD} and \eqref{eq:SNRFD} are
implemented as sums over sample points for the digitized gravitational wave
detector output.  Discrete Fourier transforms can be computed efficiently
numerically.  For that reason \eqref{eq:SNRFD} is typically far faster,
computationally.  To evaluate \eqref{eq:SNRTD} requires \order{\tmpsamps
\hoftsamps} floating point operations per filter $x_i$, where \tmpsamps\ is the
number of sample points in the filter $x_i$ and \hoftsamps\ is the number of
sample points in the data $h$.  Assuming \numtmps\ filters are required, the
total cost is \order{\numtmps \tmpsamps \hoftsamps} However, \eqref{eq:SNRFD}
requires only \order{\hoftsamps \log \hoftsamps} operations per filter assuming
transform lengths that are longer than the filter (i.e. \hoftsamps\ $>$
\tmpsamps), resulting in a total cost of \order{\numtmps \hoftsamps \log
\hoftsamps}.  In most cases \tmpsamps\ $\gg$ $\log$ \hoftsamps\ and the
computational savings by choosing the frequency domain integral form
\eqref{eq:SNRFD} is clear.  However, to take full advantage of the
computational efficiency of \eqref{eq:SNRFD} requires an acausal knowledge of
the detector data $h(t)$, which implies an inherent latency.  In contrast,
\eqref{eq:SNRTD} can be updated every time a new sample point of detector data
is taken.

@@

In order to minimize latency we propose using the time domain convolution
presented in \eqref{eq:SNRTD}.  However, because the brute force evaluation of
\eqref{eq:SNRTD} is far too costly to be useful, we will consider an
approximation to \eqref{eq:SNRTD} that can reduce substantially the cost of
\realtime\ filtering. This approximation has the form
%
% orthogonal decomposition filtering
%
\begin{equation}
\rho_i(t) \approx \sum_k^{\numslices} \sum_j^{\numsvdtmps} 
	\int_{\tau_{k}}^{\tau_{k+1}} v_{ijk} \sigma_{jk} u_{jk}\,(\tau) \, h(t-\tau) \, d \tau \label{eq:decomp}
\end{equation} 
%
%
where $u_{jk}(\tau)$ is an orthogonal basis set of filters spanning the space
of $\{x_i(\tau)\}$ and $\sigma_{jk} v_{ijk}$ is a tensor relating the filters
$u_{jk}(\tau)$ to the original filter set $\{x_i(\tau)\}$.  We claim that with
a suitable choice of filters $u_{jk}(\tau)$ one can reduce the computational
cost of \eqref{eq:SNRTD} sufficiently to feasibly search for
gravitational-waves from compact binary coalescence in \realtime.  This
requires 1) exploiting the redundancy of the template bank and 2) exploiting
the time-frequency characteristics of the binary waveforms.  We describe our
procedure for producing the decompostion in \eqref{eq:decomp} in the remainder
of this section.

@@

The first step of the orthogonal decomposition described in \eqref{eq:decomp}
is to divide the templates into \emph{time slices}.  This is a time domain
analogue to the frequency domain decomposition described in ~\cite{Marion2004,
Buskulic2010, beauville2006, beauville2008}.  A matched filter is constructed
for each time slice.  The outputs form an ensemble of partial \textsc{snr}
streams.  By linearity, these partial \SNR\ streams can be suitably time
delayed and summed to reproduce the \SNR\ of the full template.  We will show
in the next section that this, combined with the singular value docomposition,
is sufficient to enable a computationally efficient time-domain search and
furthermore is an essential part of an \earlywarning\ detection scheme.

For concreteness and simplicity, we will consider an inspiral waveform in the
quadrupole approximation, for which the time-frequency relation is
%
\begin{equation} \label{eq:fgw}
%
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{-t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5 \mu\mathrm{s}$) and $t$ is the time relative to the
coalescence of the binary~\cite{findchirppaper, kidder1992}.  Usually the
template is truncated at some prescribed time $t_0$, or equivalently frequency \fmax.
This is often chosen to correspond to the \ISCO\ defined previously. An inspiral signal
will enter the detection band at a low frequency, $f = f_\mathrm{low}$
corresponding to a time $t_\mathrm{low}$.  The template is assumed to be zero
outside the interval $[t_\mathrm{low}, t_\mathrm{0})$ and is said to have  a
duration of $t_\mathrm{0} - t_\mathrm{low}$. It is critically sampled at a
rate of $2 \fmax$.

The monotonic time-frequency relationship of \eqref{eq:fgw} allows us to choose
time slice boundaries that require substantially less bandwidth at early times
in the inspiral.  Our goal is to reduce the filtering cost of a large fraction
of the waveform by computing part of the filter integral at a lower sample
rate.  Specifically we consider here time slice boundaries with the next
highest power-of-two sample rate that critically samples the time sliced
filter.  The time slices for this template consist of the $k$ intervals $(t_k,
t_{k-1}], \dots, (t_2, t_1], (t_1, t_0]$ sampled at frequencies
$f_\mathrm{k-1}, \dots, f_1, f_0$ where $f_0 \geq 2 f_\mathrm{ISCO}$, $t_0 =
t_\mathrm{ISCO}$, $f_{k-1} \geqslant 2 f_\mathrm{low}$, and $t_k \leqslant
t_\mathrm{low}$. An example time slice design satisfying these constraints for
a $1.4 - 1.4 \, M_{\odot}$ binary is shown in table~\ref{table:time_slices}.
%
\begin{table}[h!]
\begin{minipage}[c]{0.52\textwidth}
\centering
\includegraphics{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\caption{\label{table:time_slices} Example of nearly critically sampled,
power-of-two time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from
$f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$
with a time frequency structure given by ($\ref{eq:fgw})$. $f_k$ is the sample
rate of the time slice, $(t_{k+1}, t_k]$ are the boundaries in seconds
preceeding coalescence and \slicessamps\ are the number of sample points in the
$k^{\mathrm{th}}$ filter.}
\end{table}
%

Rather than applying a unique time slice decomposition to each template
waveform, we find a decomposition that is adequate for the entire set
$\{x_i\}$.  This is facilated by choosing templates in the set $\{x_i\}$ with
similar chirp masses, $\mathcal{M}$.  The time slice decompositions of the
filters $x_i(\tau)$ lead to new filters $y_{ik}(\tau)$ satisfying
%
\begin{equation}
\label{eq:time_slices}
x_i(\tau) \approx \sum_k^{\numslices} y_{ik}(\tau)
\end{equation}
where \numslices\ is the number of time slices required.  Note that we write
the relationship as approximate since the resampling implementation may have
some inherent loss in quality.  We note that by construction these filters are
orthogonal over the index $k$ since they are disjoint in time.  In the next
section we examine how to reduce the number of filters $y_{ik}(\tau)$ via
singular value decomposition to construct a set of filters that is also
orthogonal within a given time slice.

@@

As described previously, the template banks are, by design, highly correlated.
It is possible to greatly reduce the number of filters required to achieve a
particular minimum match by designing an appropriate set of orthonormal {\em
basis templates}.  A purely numerical technique based on the application of the
singular value decomposition (\SVD) to inspiral waveforms is demonstrated
in~\cite{Cannon:2010p10398}.  Using the results of ~\cite{Cannon:2010p10398} we
establish that the filters of the previous section can be approximated to high
accuracy by the expansion in the singular value basis
%
\begin{equation}
y_{ik}(\tau) \approx \sum_j^{\numsvdtmps} \sigma_{jk} v_{ijk} u_{j}(\tau)
\label{eq:svddecomp}
\end{equation}
%
where $\sigma_{jk}$ is the $j^{\mathrm{th}}$ singular value for the
$k^{\mathrm{th}}$ time slice, $v_{ijk}$ is an orthogonal matrix for the
$k^{\mathrm{th}}$ time slice  and $u_{jk}$ is a new orthogonal basis filter set
for the $k^{\mathrm{th}}$ time slice.  The authors of \cite{Cannon:2010p10398}
showed that to high reconstruction accuracy far fewer filters are needed than
were in the original template bank. We find that when combined with the time
slice decomposition, the number of \SVD\ filters, \numsvdtmps\ is much smaller
than the original number of filters \numtmps.  We combine \eqref{eq:svddecomp}
with \eqref{eq:time_slices} to arrive at \eqref{eq:decomp}.  In the next
section we compute the expected computational cost scaling of this
decomposition and compare it with the brute force implementation in
\eqref{eq:SNRTD} and higher latency \fft\ methods.

@@

The filter bank can be implemented using finite impulse response \fir\ filters,
which are just sliding window dot products.  If there are \numtmps\ templates
of length \tmpsamps\, and the data stream contains \hoftsamps\ samples, then
applying the filter bank requires $2 \numtmps \tmpsamps \hoftsamps$ operations.

@@

It is convenient to express the computational cost of the entire filtering
procedure in floating point operations per second {\flops}.  The cost will be
the sum of the cost of the \fir\ filtering for the orthogonal filter in each
time slice plus the cost of reconstructing the original waveforms with matrix
operations and resampling.  Using the formulas in table \ref{table:flops} we
arrive at
%
\begin{equation}
%\mathrm{Computational\, cost} 
\mathrm{C.C.} \propto 2 \sum_k^{\numslices} (\slicessamps + \numtmps) \numsvdtmps f_k \nonumber
	r \sum_{k, f_k \neq f_{\fmax}}^{\numslices} \numtmps \resampsamps f_k 
\label{eq:TDcost}
\end{equation}
%
where the first sum is the total cost for filtering and reconstructing the
orthogonal filters $u_{jk}$.  The second sum is the cost of resampling the
reconstructed time slice outputs to the original sample rate {\fmax}.  It
assumes a \fir\ filter with \resampsamps\ sample points.  Note that the cost
for resampling only occurs for the downsampled time slices. The resampling cost
largely depends on {\resampsamps}.  The computational cost of \eqref{eq:TDcost}
is dominated by the highest frequency terms in the sum.  Comparing just the
highest frequency term of \eqref{eq:TDcost} with \eqref{eq:SNRTD} shows \order{
(\slicesamps{\mathrm{max}} + \numtmps) \svdtmps{\mathrm{max}} \fmax} \flops\
versus \order{\numtmps \tmpsamps \fmax} {\flops}.  The full calculation for a
particular patch of parameter space is shown in table \ref{table:comp_cost}.
\hl{FIXME we need to actually say what went into this}

@@

\editorial{It would be good to illustrate the layout of this particular
template bank: masses spanned, time slice layout \dots}
\begin{table}[!h]
\label{table:comp_cost}
\begin{center}
\input{flop_budget_example}
\end{center}
\caption{\hl{FIXME: Operation counts per sample for four different detection methods.}}
\end{table}

>>>>

In this section we describe a decomposition of the compact binary parameter
space that reduces low-latency filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multiband decomposition of the compact binary parameter space that resulted in
search with $\sim$minutes latency in \LIGO{}'s S6 and Virgo's VSR2 science runs.
We combine this with the orthogonal decomposition described
in~\cite{Cannon:2010p10398} that exploits the redundancy of the template banks.

@@

Inspiral signals are parameterized by a set of intrinsic parameters
$\bar{\theta}$ that determine the amplitude and phase evolution of the observed
binary signals. For systems where the effects of spin can be ignored, the
intrinsic parameters are the component masses of the binary, $\bar{\theta} =
(m_1, m_2)$. Searches for gravitational waves from compact binary coalescence
typically employ matched filter banks, called template banks, that discretely
sample the possible intrinsic parameters~\cite{findchirppaper}.  The filters
for the waveforms $x(t,\bar{\theta})$, known as templates, are the waveforms
weighted by the inverse detector noise amplitude spectral density, so as to
whiten them. To construct a template bank, templates are chosen with discrete
signal parameters $\theta_1,\, \theta_2,\, \cdots$, $\theta_N$ to assure a
bounded loss of \SNR~\cite{Owen:1995tm,Owen:1998dk}. That is, any possible
signal within the search space will have a cross-correlation of $\geq0.97$ with
at least one template. Such a template bank is said to have a 97\% {\em
minimum match}. Data are filtered against each template to produce an \SNR
time-series. Local peak-finding across time and templates determines detection
candidates.
%
\footnote{There are two gravitational-wave polarizations, $+$ and $\times$. A
given detector will observe a combination of these polarizations that will
largely be degenerate with an overall unknown constant phase.  This can be
maximized over by filtering for quadrature phases and taking the magnitude of
the result. For simplicity we will ignore that aspect in this work, as it is
straightforward to generalize but not necessary for understanding any of the
points that will be made.}

We will denote the $i^\mathrm{th}$ filter with parameters $\bar{\theta_i}$
as a function of time $x_i(\tau)$.
In this work we will discuss transformations to a set of filters
$\{x_i(\tau)\}$.  Some of these transformations are not useful or practical
over the entire parameter space.  For that reason we assume from here onward
that the set of filters $\{x_i(\tau)\}$ refer to a set of near-neighbor filters
that can be chosen as a subset of the full parameter space.  Several such
subsets can be chosen until all of the filters are a member of one local set.
\editorial{nvf: I don't think that the splitting of the bank is a crucial point here. Perhaps later if we discuss the up-front cost of the SVD.}

@@

The evaluation of the integrals in \eqref{eq:SNRTD} and \eqref{eq:SNRFD} are
implemented as sums over sample points for the digitized gravitational-wave
detector output.  Discrete Fourier transforms can be computed efficiently
numerically.  For that reason \eqref{eq:SNRFD} is typically far faster,
computationally.  To evaluate \eqref{eq:SNRTD} requires \order{\tmpsamps
\hoftsamps} floating point operations per filter $x_i$, where \tmpsamps\ is the
number of sample points in the filter $x_i$ and \hoftsamps\ is the number of
sample points in the data $h$.  Assuming \numtmps\ filters are required, the
total cost is \order{\numtmps \tmpsamps \hoftsamps} However, \eqref{eq:SNRFD}
requires only \order{\hoftsamps \log \hoftsamps} operations per filter assuming
transform lengths that are longer than the filter (i.e. \hoftsamps\ $>$
\tmpsamps), resulting in a total cost of \order{\numtmps \hoftsamps \log
\hoftsamps}.  In most cases \tmpsamps\ $\gg$ $\log$ \hoftsamps\ and the
computational savings by choosing the frequency-domain integral form
\eqref{eq:SNRFD} is clear.  However, to take full advantage of the
computational efficiency of \eqref{eq:SNRFD} requires an acausal knowledge of
the detector data $h(t)$, which implies an inherent latency.  In contrast,
\eqref{eq:SNRTD} can be updated every time a new sample point of detector data
is taken.

@@

In order to minimize latency we propose using the time-domain convolution
presented in \eqref{eq:SNRTD}.  However, because the brute-force evaluation of
\eqref{eq:SNRTD} is far too costly to be useful, we will consider an
approximation to \eqref{eq:SNRTD} that can reduce substantially the cost of
\realtime\ filtering. This approximation has the form
%
% orthogonal decomposition filtering
%
\begin{equation}
\rho_i(t) \approx \sum_k^{\numslices} \sum_j^{\numsvdtmps} 
	\int_{\tau_{k}}^{\tau_{k+1}} v_{ijk} \sigma_{jk} u_{jk}\,(\tau) \, h(t-\tau) \, d \tau \label{eq:decomp}
\end{equation} 
%
%
where $u_{jk}(\tau)$ is an orthogonal basis set of filters spanning the space
of $\{x_i(\tau)\}$ and $\sigma_{jk} v_{ijk}$ is a tensor relating the filters
$u_{jk}(\tau)$ to the original filter set $\{x_i(\tau)\}$.  We claim that with
a suitable choice of filters $u_{jk}(\tau)$ one can reduce the computational
cost of \eqref{eq:SNRTD} sufficiently to feasibly search for
gravitational waves from compact binary coalescence in \realtime.  This
requires 1) exploiting
the time-frequency characteristics of the binary waveforms and 2) exploiting the redundancy of the template bank. We describe our
procedure for producing the decompostion in \eqref{eq:decomp} in the remainder
of this section.

@@

The first step of the orthogonal decomposition described in \eqref{eq:decomp}
is to divide the templates into \emph{time slices}.  This is a time-domain
analogue to the frequency-domain decomposition described in ~\cite{Marion2004,
Buskulic2010, beauville2006, beauville2008}.  A matched filter is constructed
for each time slice.  The outputs form an ensemble of partial \SNR{}
streams.  By linearity, these partial \SNR\ streams can be suitably time
delayed and summed to reproduce the \SNR\ of the full template.  We will show
in the next section that this, combined with the singular value docomposition,
is sufficient to enable a computationally efficient time-domain search and
furthermore is an essential part of an \earlywarning\ detection scheme.

For concreteness and simplicity, we will consider an inspiral waveform in the
quadrupole approximation, for which the time-frequency relation is
%
\begin{equation} \label{eq:fgw}
%
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{-t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5 \umu\mathrm{s}$) and $t$ is the time relative to the
coalescence of the binary~\cite{findchirppaper, kidder1992}.  Usually the
template is truncated at some prescribed time $t_0$, or equivalently frequency \fmax.
This is often chosen to correspond to the \ISCO\ defined previously. An inspiral signal
will enter the detection band at a low frequency, $f = f_\mathrm{low}$
corresponding to a time $t_\mathrm{low}$.  The template is assumed to be zero
outside the interval $[t_\mathrm{low}, t_\mathrm{0})$ and is said to have  a
duration of $t_\mathrm{0} - t_\mathrm{low}$. It is critically sampled at a
rate of $2 \fmax$.

The monotonic time-frequency relationship of \eqref{eq:fgw} allows us to choose
time-slice boundaries that require substantially less bandwidth at early times
in the inspiral.  Our goal is to reduce the filtering cost of a large fraction
of the waveform by computing part of the filter integral at a lower sample
rate.  Specifically we consider here time slice boundaries with the next
highest power-of-two sample rate that critically samples the time sliced
filter.  The time slices for this template consist of the $k$ intervals $(t_k,
t_{k-1}], \dots, (t_2, t_1], (t_1, t_0]$ sampled at frequencies
$f_\mathrm{k-1}, \dots, f_1, f_0$ where $f_0 \geq 2 f_\mathrm{ISCO}$, $t_0 =
t_\mathrm{ISCO}$, $f_{k-1} \geqslant 2 f_\mathrm{low}$, and $t_k \leqslant
t_\mathrm{low}$. An example time-slice design satisfying these constraints for
a $1.4 - 1.4 \, M_{\odot}$ binary is shown in table~\ref{table:time_slices}.
%
\begin{table}[h!]
\begin{minipage}[c]{0.52\textwidth}
\centering
\includegraphics{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\caption{\label{table:time_slices} Example of nearly critically sampled,
power-of-two time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from
$f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$
with a time frequency structure given by ($\ref{eq:fgw})$. $f_k$ is the sample
rate of the time slice, $(t_{k+1}, t_k]$ are the boundaries in seconds
preceeding coalescence and \slicessamps\ are the number of sample points in the
$k^{\mathrm{th}}$ filter.}
\end{table}
%

Rather than applying a unique time-slice decomposition to each template
waveform, we find a decomposition that is adequate for the entire set
$\{x_i\}$.  This is facilated by choosing templates in the set $\{x_i\}$ with
similar chirp masses, $\mathcal{M}$.  The time-slice decompositions of the
filters $x_i(\tau)$ lead to new filters $y_{ik}(\tau)$ satisfying
%
\begin{equation}
\label{eq:time_slices}
x_i(\tau) \approx \sum_k^{\numslices} y_{ik}(\tau)
\end{equation}
where \numslices\ is the number of time slices required.  Note that we write
the relationship as approximate since the resampling implementation may have
some inherent loss in quality.  We note that by construction these filters are
orthogonal over the index $k$ since they are disjoint in time.  In the next
section we examine how to reduce the number of filters $y_{ik}(\tau)$ via
singular value decomposition to construct a set of filters that is also
orthogonal within a given time slice.

@@

As described previously, the template banks are, by design, highly correlated.
It is possible to greatly reduce the number of filters required to achieve a
particular minimum match by designing an appropriate set of orthonormal {\em
basis templates}.  A purely numerical technique based on the application of the
singular value decomposition (\SVD) to inspiral waveforms is demonstrated
in~\cite{Cannon:2010p10398}.  Using the results of ~\cite{Cannon:2010p10398} we
establish that the filters of the previous section can be approximated to high
accuracy by the expansion in the singular value basis
%
\begin{equation}
y_{ik}(\tau) \approx \sum_j^{\numsvdtmps} \sigma_{jk} v_{ijk} u_{j}(\tau)
\label{eq:svddecomp}
\end{equation}
%
where $\sigma_{jk}$ is the $j^{\mathrm{th}}$ singular value for the
$k^{\mathrm{th}}$ time slice, $v_{ijk}$ is an orthogonal matrix for the
$k^{\mathrm{th}}$ time slice  and $u_{jk}$ is a new orthogonal basis filter set
for the $k^{\mathrm{th}}$ time slice.  The authors of \cite{Cannon:2010p10398}
showed that to high reconstruction accuracy far fewer filters are needed than
were in the original template bank. We find that when combined with the
time-slice decomposition, the number of \SVD\ filters, \numsvdtmps\ is much smaller
than the original number of filters \numtmps.  We combine \eqref{eq:svddecomp}
with \eqref{eq:time_slices} to arrive at \eqref{eq:decomp}.  In the next
section we compute the expected computational cost scaling of this
decomposition and compare it with the brute-force implementation in
\eqref{eq:SNRTD} and higher latency \fft\ methods.

@@

The filter bank can be implemented using finite impulse response (\fir{}) filters,
which are just sliding window dot products.  If there are \numtmps\ templates
of length \tmpsamps\, and the data stream contains \hoftsamps\ samples, then
applying the filter bank requires $2 \numtmps \tmpsamps \hoftsamps$ operations.

@@

It is convenient to express the computational cost of the entire filtering
procedure in floating point operations per second {\flops}.  The cost will be
the sum of the cost of the \fir\ filtering for the orthogonal filter in each
time slice plus the cost of reconstructing the original waveforms with matrix
operations and resampling.  Using the formulas in table \ref{table:flops} we
arrive at
%
\begin{equation}
%\mathrm{Computational\, cost} 
\mathrm{C.C.} \propto 2 \sum_k^{\numslices} (\slicessamps + \numtmps) \numsvdtmps f_k \nonumber
	r \sum_{k, f_k \neq f_{\fmax}}^{\numslices} \numtmps \resampsamps f_k 
\label{eq:TDcost}
\end{equation}
%
where the first sum is the total cost for filtering and reconstructing the
orthogonal filters $u_{jk}$.  The second sum is the cost of resampling the
reconstructed time-slice outputs to the original sample rate {\fmax}.  It
assumes a \fir\ filter with \resampsamps\ sample points.  Note that the cost
for resampling only occurs for the downsampled time slices. The resampling cost
largely depends on {\resampsamps}.  The computational cost of \eqref{eq:TDcost}
is dominated by the highest frequency terms in the sum.  Comparing just the
highest frequency term of \eqref{eq:TDcost} with \eqref{eq:SNRTD} shows \order{
(\slicesamps{\mathrm{max}} + \numtmps) \svdtmps{\mathrm{max}} \fmax} \flops\
versus \order{\numtmps \tmpsamps \fmax} {\flops}.  The full calculation for a
particular patch of parameter space is shown in table \ref{table:comp_cost}.
\hl{FIXME we need to actually say what went into this}

@@

\editorial{It would be good to illustrate the layout of this particular
template bank: masses spanned, time-slice layout \dots}
\begin{table}[!h]
\label{table:comp_cost}
\begin{center}
\input{flop_budget_example}
\end{center}
\caption{\hl{FIXME: Operation counts per sample for four different detection methods.}}
\end{table}
